The Weakly Nonlinear Saturation of the Magnetorotational Instability

### 2014-09-17 paper/mri_paper.tex
more

We develop this investigation with the goal of providing a rigorous analytical framework that will be used to make predictions for laboratory MRI experiments. In recent years much progress has been made toward the goal of observing MRI in a laboratory setting, although no experiment has yet yielded a confirmed detection. Of most immediate interest to this work is the Princeton Plasma Physics Laboratory (PPPL) MRI experiment, which has a design roughly analogous to the set-up investigated here (see \S\ref{setup} for further details). The Madison plasma dynamo experiment is a multi-purpose plasma device which may be a vehicle for MRI studies in the near future (\citei{Cooper:2013to}). The helical MRI, an instability related to the standard MRI but easier to excite in the laboratory, is the focus of the Potsdam Rossendorf Magnetic Instability Experiment (PROMISE) in Dresden. PROMISE has already detected helical MRI in the laboratory (e.g. \citei{Stefani:2006iv}), and we discuss possible applications of our framework to the helical MRI in \S\ref{discussion}.

@@

We compute the solutions to our derived equations pseudospectrally, using Dedalus. One advantage to using Dedalus for these computations is that the code can compute the evolution of the fully nonlinear MRI system on the same grid.

@@

@@

\bibitem[{Balbus \& Hawley(1991)}]{Balbus:1991vs}
Balbus, S A and Hawley, J F, The Astrophysical Journal, 376, 214

@@

\bibitem[{Cooper et al.(2014)}]{Cooper:2013to}
Cooper, C M, Wallace, J, Brookhart, M, Clark, M, Collins, C, Ding, W X, Flanagan, K, Khalzov, I, Li, Y, Milhone, J, Nornberg, M, Nonn, P, Weisberg, D, Whyte, D G, Zweibel, E, Forest, C B, Physics of Plasmas, 21, 013505

\bibitem[{Ji et al.(2001)}]{Ji:2001kd}
Ji, H, Goodman, J, Kageyama, A, Mon. Not. R. Astron. Soc., 325, 1

\bibitem[{Schartman et al.(2009)}]{Schartman:2009df}
Schartman, E, Ji, H, Burin, M J, Rev Sci Instrum, 80, 024501

\bibitem[{Shakura \& Sunyaev(1973)}]{Shakura:1973wg}
Shakura, N I and Sunyaev, R A, Astronomy \& Astrophysics, 24, 337

\bibitem[{Stefani et al.(2006)}]{Stefani:2006iv}
Stefani, F, Gundrum, T, Gerbeth, G, RŸdiger, G, Schultz, M, Szklarski, J, Hollerbach, R, Phys Rev Lett, 97, 184502

\bibitem[{Umurhan et al.(2007a)}]{Umurhan:2007dz}
Umurhan, O M, Regev, O, Menou, K, Phys. Rev. Letters, 98, 034501

\bibitem[{Umurhan et al.(2007b)}]{Umurhan:2007hs}
Umurhan, O M, Regev, O, Menou, K, Phys. Rev. E, 76, 036310

>>>>

We develop this investigation with the goal of providing a rigorous analytical framework that will be used to make predictions for laboratory MRI experiments. Observing MRI in a laboratory setting would open a new avenue for studying MRI dynamics and saturation. In recent years much experimental progress has been made, although no experiment has yet yielded a confirmed detection. Of most immediate interest to this work is the Princeton Plasma Physics Laboratory (PPPL) MRI experiment, which has a design roughly analogous to the set-up investigated here (see \S\ref{setup} for further details). The Madison plasma dynamo experiment is a multi-purpose plasma device which may be a vehicle for MRI studies in the near future (\citei{Cooper:2013to}). The helical MRI, an instability related to the standard MRI but easier to excite in the laboratory, is the focus of the Potsdam Rossendorf Magnetic Instability Experiment (PROMISE) in Dresden. PROMISE has already detected helical MRI in the laboratory (e.g. \citei{Stefani:2006iv}), and we discuss possible applications of our framework to the helical MRI in \S\ref{discussion}. A spherical Couette experiment claimed to have detected MRI in 2004 (\citeauthor{Sisan:2004ig}), but most likely detected unrelated MHD instabilities instead (\citei{Hollerbach:2009ig}; \citei{Gissinger:2011td}).

@@

We compute the solutions to our derived equations pseudospectrally, using Dedalus. One advantage to using Dedalus for these computations is that the code can compute the evolution of the fully nonlinear MRI system on the same Fourier/Chebyshev grid. This allows us to directly compare simulation and theory.

@@

The present configuration, that of magnetized Taylor Couette flow with a purely axial magnetic field, represents the initial conditions for the standard MRI. The helical MRI is a related, but separate, MHD instability.

@@

\bibitem[{Balbus \& Hawley(1991)}]{Balbus:1991vs}
Balbus, S A and Hawley, J F, 1991, The Astrophysical Journal, 376, 214

@@

\bibitem[{Cooper et al.(2014)}]{Cooper:2013to}
Cooper, C M, Wallace, J, Brookhart, M, Clark, M, Collins, C, Ding, W X, Flanagan, K, Khalzov, I, Li, Y, Milhone, J, Nornberg, M, Nonn, P, Weisberg, D, Whyte, D G, Zweibel, E, Forest, C B, Physics of Plasmas, 2013, 21, 013505

\bibitem[{Gissinger et al.(2011)}]{Gissinger:2011td}
Gissinger, C, Ji, H, Goodman, J, Phys. Rev. E, 2011, 84, 026308

\bibitem[{Hollerback(2009)}]{Hollerbach:2009ig}
Hollerback, R, Proc. R. Soc. A, 2009, 465, 2107

\bibitem[{Ji et al.(2001)}]{Ji:2001kd}
Ji, H, Goodman, J, Kageyama, A, Mon. Not. R. Astron. Soc., 2001, 325, 1

\bibitem[{Schartman et al.(2009)}]{Schartman:2009df}
Schartman, E, Ji, H, Burin, M J, Rev Sci Instrum, 2009, 80, 024501

\bibitem[{Shakura \& Sunyaev(1973)}]{Shakura:1973wg}
Shakura, N I and Sunyaev, R A, Astronomy \& Astrophysics, 1973, 24, 337

\bibitem[{Sisan et al.(2004)}]{Sisan:2004ig}
Sisan, D R, Mujica, N, Tillotson, W A, Huang Y-M, Dorland, W, Hassam, A B, Antonsen, T M, Lathrop, D P, Phys Rev Lett, 2004, 93, 11

\bibitem[{Stefani et al.(2006)}]{Stefani:2006iv}
Stefani, F, Gundrum, T, Gerbeth, G, RŸdiger, G, Schultz, M, Szklarski, J, Hollerbach, R, Phys Rev Lett, 2006, 97, 184502

\bibitem[{Umurhan et al.(2007a)}]{Umurhan:2007dz}
Umurhan, O M, Regev, O, Menou, K, 2007, Phys. Rev. Letters, 98, 034501

\bibitem[{Umurhan et al.(2007b)}]{Umurhan:2007hs}
Umurhan, O M, Regev, O, Menou, K, 2007, Phys. Rev. E, 76, 036310
